Physics-based protein structure prediction and design using the confinement method

### 2012-09-09 Paper/paper.tex
Reworked Orban section

%\documentclass[preprint,preprintnumbers,amsmath,amssymb]{revtex4}
\documentclass[12pt]{article}

%\usepackage{fontspec} 
%\setmainfont{TeX Gyre Pagella}
\usepackage{amsmath}
\usepackage[nofiglist,notablist]{endfloat}
\usepackage[usenames,dvipsnames]{color}
\usepackage{color}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{palatino}
\usepackage[activate={true,nocompatibility},final]{microtype}
\usepackage[super,sort&compress]{natbib}
\pagenumbering{arabic}
\parskip = 0.08in \parindent = 0.0in

% Custom macros for author comments
\newcommand{\Alberto}[1]{\color{ForestGreen}Alberto: #1 \normalcolor }
\newcommand{\Justin}[1]{\color{blue}Justin: #1 \normalcolor}
\newcommand{\Arijit}[1]{\color{yellow}Arijit: #1 \normalcolor}
\newcommand{\Ken}[1]{\color{red}Ken: #1 \normalcolor}

\author{Arijit~Roy$^1$}
\author{Justin~L.~MacCallum$^1$}
\author{Alberto~Perez$^1$}
\author{Ken~A.~Dill$^1$}
\affil{$^1$Laufer Center for Physical and Quantitative Biology\\
    Stony Brook University\\
    Stony Brook, NY 11794-5252.}

\title{Ranking protein structures using confinement free energy method}

@@

\maketitle
\begin{abstract}
%How to cite something in paper.bib Abstract \citep{CASP9}

@@

%of protein can be obtained to atomic resolution by X-ray crystallography or NMR. Unfortunately, the sheer amount
%of time and economic investment for structure determination cannot keep up with the rate at which
%proteins of unknown structure are sequenced. Therefore, computational models can be a good way to bridge the
%increasing gap between sequence and structure. In this respect CASP (Critical Assessment of Structure Prediction) has played a 
%crucial role which is a community wide experiments for the protein structure prediction.

The confinement method has been described in details in ref. by Tyka et.al. and Cecchini et. al. Here we briefly describe the methodology.
We compute free energy $\Delta G_{AB}$ between A and B conformation of the protein. For this purpose we use a thermodynamic cycle.
\begin{enumerate}

\section{Result}

As a proof of priciple we first reproduced the examples from previous work where this method has been applied to a 
16 amino acid residue peptide, $\beta$ hairpin from protein G popularly known as BHP to calculate the free energy difference 
between two of its conformation.

\subsection{BHP}

First we calculated the free energy difference between two conformation of the polypeptide $\beta$ hairpin from protein G.  
This peptide contain 16 residue. One of the conformation is the 
native structure, has a two stranded $\beta$ sheet known as bhp1 and the other one is known as bhp3 which has a three stranded $\beta$ sheet.
The free energy difference between them was calculated using 50 independent MD simulation each of which has a lengh of 4 $\mu$ second totaling
200 $\mu$ second. The free energy calculated was 1.8 Kcal/mol at 360 K. On the other hand using confinement method, we found that the 
free energy difference between bhp2 and bhp1 is 1.67 $\pm$  .

\subsection{Protein with similar sequence but different fold}

In general, the protein structure with similar kind of sequences tends to have similar 3D structure and are usually assumed to be 
homologous. This idea is the basis of comparative modeling and fold recognition in protein structure prediction.
Although there are examples in literature that protein can undergo changes in their fold due to cleavage of a peptide bond 
or due to few changes in amino acid sequence but a more dramatic event happen for the case of two small engineered protein 
consisting 56 amino acid residues from domain of
protein G, known as protein GA and GB. The protein GA is serum binding domain with $3 \alpha$ fold and the protein GB is an 
IgG-binding protein with $4 \beta + \alpha$ fold. Two pairs of such protein are engineered by step by step mutation until they 
change fold. One pair that have 88 \% similar sequence are known as GA88 and GB88 and another pair with 95 \% similar sequence 
are known as GA95 and GB95. There are differences in 7 and 3 positions with 88 \% and 95 \% similar sequence respectively and also can 
be visible from Figure \ref{fig:G88} and Figure \ref{fig:G95}. The difference between GA95 and GB95 is only at residue positions 20, 
30 and 45. While GA95 have amino acid residues Leu , Ile and Leu at positions 20, 30 and 45, in case of GB95 the amino acid residues in
these positions are Ala, Phe and Tyr. On the other hand there are amino acid residues Gly-24, Ile-25, Ile-30, Ile-33,Leu-45,Ile-49 and 
Leu-50 in GA88 and in GB88 in similar positions there are residues Ala-24, Thr-25, Phe-30, Tyr-33, Tyr-45, Thr-49 and Lys-50. 
We have calculated the free energy difference between GA and GB with 
both 88 \% and 95 \% sequence similarity. In order to do such calculation, we have first taken the sequence of GA95 and generated the both 
GA ($3 \alpha$) fold and GB $4 \beta + \alpha$ fold with the same sequence using a specialized restraint based replica exchange 
molecular dynamic simulation. The similar procedure was followed to generate the structure with the sequence of GB95, GA88 and GB88.
% write why we used this %
In the process we generated total of four pairs of structure and we also checked that the structures were withing 2 \AA rmsd of the
NMR generated structure in all the cases. We expect the free energy should be lowest in case of sequence with the native 
conformation. The result is presented in Table~\protect\ref{tab:GA_GB}.  
In all cases the sequence with the native structure has the lowest free energy value. Also, the low free energy difference support
the hypothesis that due to such low free energy difference these proteins can switch their fold.

\subsection{Application on different targets of CASP9 experiments}

Next we have applied the confinement method to different targets of CASP(Critical Assessment of Structure Prediction) 9 experiments. 
CASP is a world wide competition aiming at finding the state of the art methodologies in protein folding. Groups, both from knowledge 
based methods and physics based methods participate in this blind competition and for each sequence target they can submit 
five models. The predictors were also asked to calculate their best model among the set of five they submitted. However, in many cases
predictors were unable to predict/rank their models. Using the confinement method, we aim to rank protein structures using free 
energy as a scale. In recent CASP experiments Global distance test scose (GDT\_TS) method is used for final analysis of different
submitted structures with the NMR/crystallographically solved protein native structure. GDT\_TS represent the average 
percentage of residues that are in close proximity in two structures optimally superimposed
using four different distance cutoff. Although GDT\_TS alone is not a reliable measure for the difficult modeling cases, it has 
been well accepted method in recent years among the community. We have also compared our free energy result with the 
GDT\_TS score for different cases we studied.

%The GDTTS value of a model is determined as follows. A large sample of possible structure superpositions of the model on the corresponding
%experimental structure is generated by superposing all sets of three, five, and seven consecutive C$\alpha$ atoms along the backbone (each
%peptide segment provides one superposition). Each of these initial superpositions is iteratively extended, including all residue pairs under
%a specified threshold in the next iteration, and continuing until there is no change in included residues. The procedure is carried out using
%thresholds of 1, 2, 4, and 8 $\AA$, and the final superposition that includes the maximum number of residues is selected for each threshold.

@@

\subsubsection{T0605}
This is a 72 amino acid residue protein but for the final calculation only the residues 18-66 amino acids were kept for the final analysis.
The best performing group was "Baker" for this target. We have carried out our calculation with the model first, second and fifth for the
final calculation. The rest two model had closer GDT\_TS value compared to the previously mentioned models and thus they were omitted from 
furthe calculation. The result is presented in Table~\protect\ref{tab:T0605}. (This is a case where we can demand the confinement 
method is better than GDT\_TS).

@@

\begin{table}
\caption{The comparison of free energy values of proteins with 95 \% and 88 \% similar sequences in $3 \alpha$ and $4 \beta + \alpha$ fold.}
\label{tab:GA_GB}
\begin{center}
\begin{tabular}{l l l}\hline
Changes in   &     $\Delta G$ (Kcal/Mol) &  $\Delta G$ (Kcal/Mol) \\
 Sequences   &     ($3 \alpha$ fold)       & ($4 \beta + \alpha$ fold) \\ \hline
L20I30L45    &     $0.000$            &  $2.901 \pm 0.475$   \\
A20F30Y45    &     $4.365 \pm 0.460$ &  $0.000$    \\
G24I25I30I33 &     $0.000$            &  $ 3.941 \pm 0.517$ \\
L45I49L50    &                        &              \\
A24T25F30Y33 &     $5.576 \pm 0.464$  &  $0.000$     \\
Y45T49K50    &                        &             \\ \hline
\end{tabular}
\end{center}
\end{table}

@@

>>>>

\documentclass[12pt]{article}

\usepackage{amsmath}
\usepackage[nofiglist,notablist]{endfloat}
\usepackage[usenames,dvipsnames]{color}
\usepackage{color}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{palatino}
\usepackage[activate={true,nocompatibility},final]{microtype}
\usepackage[super,sort&compress]{natbib}
\pagenumbering{arabic}
\parskip = 0.08in \parindent = 0.0in

% Custom macros for author comments
\newcommand{\Alberto}[1]{\color{ForestGreen}#1\normalcolor }
\newcommand{\Justin}[1]{\color{blue}#1\normalcolor}
\newcommand{\Arijit}[1]{\color{yellow}#1\normalcolor}
\newcommand{\Ken}[1]{\color{red}#1\normalcolor}

\author{Arijit~Roy}
\author{Justin~L.~MacCallum}
\author{Alberto~Perez}
\author{Ken~A.~Dill}
\affil{Laufer Center for Physical and Quantitative Biology\\
    Stony Brook University\\
    Stony Brook, NY 11794-5252.}

\title{Physics-based protein structure prediction and design using the confinement method}

@@

\maketitle

\begin{abstract}

@@

%of protein can be obtained to atomic resolution by X-ray crystallography or NMR. Unfortunately, the
%sheer amount of time and economic investment for structure determination cannot keep up with the
%rate at which proteins of unknown structure are sequenced. Therefore, computational models can be a
%good way to bridge the increasing gap between sequence and structure. In this respect CASP
%(Critical Assessment of Structure Prediction) has played a crucial role which is a community wide
%experiments for the protein structure prediction.

\section{Results and Discussion}

\subsection{The confinement method produces correct results in control experiments}

As a first step, we performed several control experiments to verify that our implementation of the
confinement method produces results compatible with previous calculations reported in the
literature.

The method has previously been applied to a 16 amino acid residue $\beta$-hairpin from
protein G, known as BHP[ref]. We calculated the free energy difference between two different
conformations of the peptide: (1) the native conformation, called bhp1, with a two-stranded
$\beta$-sheet; and (2) a conformation, called bhp3, which has a three-stranded $\beta$-sheet.
Analysis of long (4 $\mu$s) equilibrium simulations[ref] shows that bhp1 is the more favorable
configuration by 1.8 kcal/mol. Using the confinement method, we obtain a value 1.7 kcal/mol, which
is in good agreement with the equilibrium simulations and with previous calculations using the
confinement method[ref] (see Supporting Information for further details).

\Justin{Arijit, Al: did we do any other control experiments on small peptides that we should report
here?}

Previous applications of the confiment method have focused on relatively short peptides, up to X
residues in length \Justin{Arijit: fill in X}. In this work we apply the confinement method to
larger proteins. One of the  applications we explore in the present work is the re-scoring, or
metaprediction, of structures submitted during the Critical Assessment of Structure Prediction
(CASP) experiment (described in detail later). We computed the relative free energies of predictions
submitted during CASP9 and 10, with the expectation that the most native-like prediction will have
the lowest free energy. As a control, for several targets we also calculated the relative free
energy of the experimental structure (once it was available from the PDB). As
Table~\ref{table:casp_control} shows, in X of Y cases, the confinement method correctly assigns a
lower free energy to the experimentally determined structure than to any of the decoys. Although
differentiating between experimentally determined structures and computer generated predictions is
not a stringent test of a scoring method---for example, examining the residue-residue packing can
easily distinguish between the two [ref Rosetta Holes]---the results none the less serve as a useful
"sanity check".

\begin{table}
\label{table:casp_control}

\begin{center}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier & $\Delta\Delta G_{native \to best decoy}$ (kcal/mol) \\ \hline
\end{tabular}
\end{center}

\caption{The confinement method assigns a more favorable free energy to the experimentally
determined structure than to computer-generated predictions. For each target, we examined as many as
five predictions submitted by CASP participants. Positive $\Delta\Delta G$ values indicate that the
experimental structure is predicted to be more favorable than any of the decoys.}

\end{table}

\subsection{The confinement method correctly predicts the structural preferences of chameleon
sequences}

In general, proteins with similar sequences tend to have similar structures. This idea is the basis
of comparative modeling and fold recognition in protein structure prediction. There are, however,
examples---often referred to as chameleon sequences---of proteins with similar sequences that have
remarkably different structures. Orban and co-workers have designed a series of 56-residue proteins
(based on Protein G) that adopt one of two different folds depending on small changes in sequence
(see Figure~\ref{fig:orban}). Sequences that adopt a mixed alpha/beta structure similar to Protein G
are denoted as ``GB'', while sequences that form a three-helix bundle are denoted as ``GA''. One
pair of sequences (GA88/GB88) are 88 percent identical and differ in seven positions. Another pair
(GA95/GA95) are 95 percent identical and differ in three positions. The last pair (GA98/GB98) differ
only in a single tyrosine to alanine mutation.

The fact such small changes in sequence can lead to such dramatic changes in structure is rather
remarkable. Accurately predicting the structural preferences of these structures presents a serious
challenge for computational methods [ref].

We initially approached this problem by making a model of each sequence with the same backbone
structure as its partner chameleon sequence. For example, we took the sequence of GA88 and built a
model with the same overall structure as GB88. We then used the confinement method to assess the
free energy difference between the experimentally determined structure of GA88 and the model (with
the GA88 sequence and the GB88 structure). The confinement method was able to predict the
conformational preferences correctly for all six sequences (data not shown). There is, however, a
serious problem with this analysis: we are comparing an experimentally determined structure with a
computer generated model. It might be possible that we were able to make correct predictions simply
because artefacts of our modeling procedure always lead to the model having a higher free energy
than the experimental structure.

To avoid this potential problem, we instead computed the relative free energy two different
structural models for each sequence. One model is based on the GA structure and the other on the GB
structure (see Supporting Information for details on the modeling procedure). This is a much more
realistic test of the confinement method's ability to accurately calculate relative free energies.

The results of these calculations are presented in Figure~\ref{fig:orban}. The confinement method
identifies the correct structure for all six sequences. The free energy differences range from
around 2.9 to 5.7 kcal/mol, which is consistent with small changes in sequences and with estimates
made by Orban and co-workers [ref].

\begin{figure}
\label{fig:orban}

\caption{The confinement method correctly predicts the structural preferences of six chameleon
sequences. (A) The six sequences used in this study. (B) Each sequence adopts either a Protein
G-like fold (denoted GB) or a three-helix bundle fold (denoted GA). The relative free energies of
the two folds are reported for each sequence.}

\Justin{Arijit: we need to make this figure, which we've discussed before and move the data from the
following table into the figure.}
\end{figure}

%\begin{table}
%\caption{The confinement method correctly reproduces the structural preferences of a series of
%designed peptides. Each sequence adopts either a Protein G-like fold ($4 \beta + \alpha$) or forms a
%helix bundle ($3 \alpha$).}
%\label{table:orban}
%\begin{center}
%\begin{tabular}{l l l}\hline
%Changes in & $\Delta G$ (kcal/mol)    & $\Delta G$ (kcal/mol)     \\
 %Sequences & ($3 \alpha$ fold)        & ($4 \beta + \alpha$ fold) \\ \hline
%GA95       & $0.000$                  & $2.901 \pm 0.475$         \\
%GB95       & $4.365 \pm 0.460$        & $0.000$                   \\
%GA88       & $0.000$                  & $ 3.941 \pm 0.517$        \\
%GB88       & $5.576 \pm 0.464$        & $0.000$                   \\
%GA98       & \Justin{Arijit: fill in} &                           \\ \hline
%GB98       &                          &                           \\ \hline
%\end{tabular}
%\end{center}
%\end{table}

\subsection{The confinement method correctly identifies the most native-like predictions from a
subset of CASP predictions}

Next, we have applied the confinement method to several targets from the CASP9 and CASP10
experiments. CASP is a world-wide competition aiming at finding the state of the art methodologies
in protein folding. Groups, both from knowledge based methods and physics based methods participate
in this blind competition and for each sequence target they can submit five models. The predictors
were also asked to calculate their best model among the set of five they submitted. However, in many
cases predictors were unable to predict/rank their models. Using the confinement method, we aim to
rank protein structures using free energy as a scale. In recent CASP experiments Global distance
test scose (GDT\_TS) method is used for final analysis of different submitted structures with the
NMR/crystallographically solved protein native structure. GDT\_TS represent the average percentage
of residues that are in close proximity in two structures optimally superimposed using four
different distance cutoff. Although GDT\_TS alone is not a reliable measure for the difficult
modeling cases, it has been well accepted method in recent years among the community. We have also
compared our free energy result with the GDT\_TS score for different cases we studied.

%The GDTTS value of a model is determined as follows. A large sample of possible structure
%superpositions of the model on the corresponding experimental structure is generated by superposing
%all sets of three, five, and seven consecutive C$\alpha$ atoms along the backbone (each peptide
%segment provides one superposition). Each of these initial superpositions is iteratively extended,
%including all residue pairs under a specified threshold in the next iteration, and continuing until
%there is no change in included residues. The procedure is carried out using thresholds of 1, 2, 4,
%and 8 $\AA$, and the final superposition that includes the maximum number of residues is selected
%for each threshold.

@@

\subsubsection{T0605}

This is a 72 amino acid residue protein but for the final calculation only the residues 18-66 amino acids were kept for the final analysis.

The best performing group was "Baker" for this target. We have carried out our calculation with the model first, second and fifth for the
final calculation. The rest two model had closer GDT\_TS value compared to the previously mentioned models and thus they were omitted from
furthe calculation. The result is presented in Table~\protect\ref{tab:T0605}. (This is a case where we can demand the confinement
method is better than GDT\_TS).

@@

@@

The confinement method has been described in details in ref. by Tyka et.al. and Cecchini et. al.
Here we briefly describe the methodology.  We compute free energy $\Delta G_{AB}$ between A and B
conformation of the protein. For this purpose we use a thermodynamic cycle.

\begin{enumerate}
